# Dichotomous Rasch Models {#Dich_Rasch_model}

## Call Back
$$Pr(Xi=1|θs,δi)=exp(θs−δi)1+exp(θs−δi)$$

## R-Lab: Running Dichotomous Rasch Model in R

### Prepare the R Packages for analysis
In this session, we use "TAM" package to do the Rasch Analysis. TAM is actually the abbreviation for *"Test Analysis Modules"*. Different from the *Winstep* software, it applies marginal maximum likelihood estimation (MMLE) instead of joint maximum likelihood estimation (JMLE).
Note: Because of the different way of the estimation, the result might be slightly different with your *Winstep* outcomes.

```{r message=FALSE, warning=FALSE}
# Load the R package that we need for this analysis
library("TAM") # For Dichotomous Rasch Analysis
library("WrightMap") # For plotting the variable map
library("Hmisc") # For descriptive data analysis
```

### Import the data & Running descriptive analysis
```{r message=FALSE}
library(readr) # Import the data from your computer
transreas <- read_csv("transreas.csv")
```
Then,we run a descriptive analysis on our data.
```{r}
# Use the summary() function to overview the data
summary(transreas)
# From the result, we can see there is no missing data for each variable. And we can also get a general idea on the range of the grades (from 2 to 6), Min, Max, Median for each task. 
# You can also use describe() function as an alternative approach
describe(transreas)
```

### Runing the Dichotomous Rasch Model
For running the dichotomous Rasch Model using TAM package, we only need students' scores. So the first and second column of our data are not needed. 
```{r}
# Trim the data
Di_Rash_data <- transreas[,c(-1,-2)]
head(Di_Rash_data) # Take a look
```

```{r results='hide'}
# Running the Rasch Dichotomous Model
Di_Rasch_model <- tam(Di_Rash_data)
```

### Overall Model Summary

```{r}
# Check the summary
summary(Di_Rasch_model)
# Plot the variable-Map
IRT.WrightMap(Di_Rasch_model,show.thr.lab=FALSE)
```

### Item Parameters
First of all, let's pool out the item parameters from your model.
```{r}
difficulty <- Di_Rasch_model$xsi
head(difficulty) 
```

- The 'xis' column denotes the item difficulty in a logit scale. And 'se.xsi' is the standard error for each item. The standard error tells us whether item difficulties may overlap or not.

- Since the `xsi` indicates the item difficulty, the higher the value refer to the harder the item. For instance, item 9 is the hardest item ( `xsi` = 1.00), whereas the item 6 is the easiest item (`xsi` = -4.07).

We can also visualize the item difficulty by using a simple histogram
```{r}
hist(difficulty$xsi,breaks=10) 
```

Now let's calculate the Item fit statistics
```{r}
Di_Item_fit <- tam.fit(Di_Rasch_model) 
head(Di_Item_fit)
```

### Person Parameters
Get the person ability by using `tam.wle` function

```{r}
Person_ability <- tam.wle(Di_Rasch_model)
# View(Person_ability)
# In the data frame above, the `theta` is the person's ability measure in a logit scale.
mean(Person_ability$theta)
``` 

Visualize the Person ability by using a simple histogram
```{r}
hist(Person_ability$theta)
```


Calculate the Person fit statistics
```{r}
Di_Person_fit <- tam.personfit(Di_Rasch_model) 
head(Di_Person_fit)
```

## Reference

[Li, Y. Using the open-source statistical language R to analyze the dichotomous Rasch model. Behavior Research Methods 38, 532–541 (2006). https://doi.org/10.3758/BF03192809](https://link.springer.com/content/pdf/10.3758/BF03192809.pdf)


